---
title: "DataScience03"
author: "Jim Stearns, in collaboration with Bethene Britt and Lucas Pedroza"
date: "25 October 2014"
output: html_document
---

In-Class Assignment, Week 3
===
UW PCE Data Science, Introduction, Fall 2014. Instructor: Dr. Ernst Henle

1. Create an R-file called DataScience03.R
====

Creating an R Markdown file: DataScience.Rmd. This file.

2. Download the mammographic_masses.data
```{r}
url <- "http://archive.ics.uci.edu/ml/machine-learning-databases/mammographic-masses/mammographic_masses.data"
# Avoid conversion of columns to factors.
mamm <- read.csv(url, header=F, stringsAsFactors=FALSE)

head(mamm)
```

3. Assign a vector of real names to the data frame
====
Real names are here 
http://archive.ics.uci.edu/ml/machine-learning-databases/mammographic-masses/mammographic_masses.names

Copied from Section 7 of link above:
Attribute Information:

1. BI-RADS assessment: 1 to 5 (ordinal)  
2. Age: patient's age in years (integer)
3. Shape: mass shape: round=1 oval=2 lobular=3 irregular=4 (nominal)
4. Margin: mass margin: circumscribed=1 microlobulated=2 obscured=3 ill-defined=4 spiculated=5 (nominal)
5. Density: mass density high=1 iso=2 low=3 fat-containing=4 (ordinal)
6. Severity: benign=0 or malignant=1 (binominal)

```{r}
names(mamm) <- c("BI_RADS", "Age", "Shape", "Margin", "Density", "Severity")
str(mamm)
#hist(mamm$BI_RADS) Error: 'x' must be numeric
sapply(mamm, is) # No "?" in final (result) column because any ? in predictors will yield result of ?
```
4. Make sure all columns are not factors
====

Done above by using parameter stringAsFactors=False in read.csv.

5. Replace "?" with NA
====

From class: "Data is big. Throw away ill-formed data. If you throw 90% away, OK, then think about it. But if you've got 5M observations and only need 50K, tossing 95% is OK."

```{r}
nrow_before <- nrow(mamm)
print(sprintf("Nrows before NA removal=%d", nrow_before))

mamm_as_matrix <- sapply(mamm, as.numeric)

# Get the count of rows with one or more NA cells. 
# Note: Check for NA (Not Available), not NaN (Not a Number)
bad_rows <- mamm_as_matrix[!complete.cases(mamm_as_matrix),]
print(sprintf("Number of rows with at least one NA, according to complete.cases() = %d",
              nrow(bad_rows)))

# Strip rows with a NA
mamm_as_matrix <- na.omit(mamm_as_matrix)

mamm_no_nas <- data.frame(mamm_as_matrix)
str(mamm_no_nas)
nrow_after <- nrow(mamm_no_nas) 
nrows_removed <- nrow_before - nrow_after
nrows_removed
```
There were `r nrows_removed` rows with at least one NA removed.

6. Determine obvious outliers using histograms, summary functions, and the data dictionary.
====

Check all column values for outliers outside the range of legal values for that column.
Don't toss observations with outlier(s); just bring the outlier into range.

Specify the ranges for all the columns in a dataframe:
```{r}
# Col1 (BI_RADS): 1..5
# Col2 (Age): 0..150 (range picked by me)
# Col3: Shape: 1..4
# Col4: Margin: 1..5
# Col5: Density: 1..4
# Col6: Severity: 0..1

ranges <- data.frame(matrix(ncol=6, nrow=2))
colnames(ranges) <- colnames(mamm)
rownames(ranges) <- c("lower", "upper")
ranges["lower", "BI_RADS"] <- 1
ranges["upper", "BI_RADS"] <- 5

ranges["lower", "Age"] <- 0
ranges["upper", "Age"] <- 150

ranges["lower", "Shape"] <- 1
ranges["upper", "Shape"] <- 4

ranges["lower", "Margin"] <- 1
ranges["upper", "Margin"] <- 5

ranges["lower", "Density"] <- 1
ranges["upper", "Density"] <- 4

ranges["lower", "Severity"] <- 0
ranges["upper", "Severity"] <- 1

ranges
```
Now check each column for outlier values below or above the legal range
```{r}
check_bound <- function(mamm, colidx, bound_type, bound_value) {
    
    col_values <- mamm[,colidx]
    if (bound_type == "upper") {
        col_vector <- (col_values > bound_value)
    } else {
        col_vector <- (col_values < bound_value)
    }
    noutliers = sum(col_vector == TRUE)
    if (is.null(noutliers)) { 
        noutliers = 0 
    }
    if (noutliers > 0) {
        mamm[col_vector == TRUE, colidx] <- bound_value
        print(sprintf("Changed these %d %s bound outliers for %s to value %d:", 
                      noutliers, bound_type, col_name, bound_value))
        print(mamm[col_vector == TRUE, colidx])
    }
    else {
         print(sprintf("No %s outliers for %s", bound_type, col_name))
    }
    
    # Function parameters in R are pass by value, not pass by reference.
    # Implication: changes to mamm parameter are not seen in calling environment,
    # unless mamm is also returned.
    return(list("noutliers"= noutliers, "mamm" = mamm))
}

for (colidx in 1:ncol(ranges)) {
    col_name <- colnames(ranges)[colidx]
    
    # Initial histogram before any range bounding
    hist(mamm_no_nas[ , colidx], breaks=100, xlab="",
         main=sprintf("%s before any range bounding", col_name))
    
    ##
    # Lower outlier check
    ##
    bound_type = "lower"
    bound_value <- ranges[bound_type, colidx]
    check_bound_return_list = check_bound(
        mamm_no_nas, colidx, bound_type, bound_value)
    # Unpack the pair of values returned in a list by check_bound
    n_lower_outliers <- check_bound_return_list$noutliers
    mamm_no_nas <- check_bound_return_list$mamm
    
    ##
    # Upper outlier check
    ##
    bound_type <- "upper"
    bound_value <- ranges[bound_type, colidx]
    check_bound_return_list = check_bound(
        mamm_no_nas, colidx, bound_type, bound_value)
    # Unpack the pair of values returned in a list by check_bound
    n_upper_outliers <- check_bound_return_list$noutliers
    mamm_no_nas <- check_bound_return_list$mamm
        
    # If any outliers were found and fixed, show a new histogram
    if (n_lower_outliers + n_upper_outliers > 0) {
        hist(mamm_no_nas[ , colidx], breaks=100, xlab="",
         main=sprintf("%s after %d outliers moved into range", 
                      col_name, n_lower_outliers + n_upper_outliers))
    }
}

mamm_clean <- mamm_no_nas
```

7. Write the data frame to a csv file called mamm.csv
====
Skip writing a number as a row name.
```{r}
write.csv(mamm_clean, file="mamm.csv", row.names=FALSE)